The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-03-31 paper.qmd
tidied section 2, and Future work - look for XXX on questions/comments

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---

@@

# What is Uncertainty?
## Colloquial definitions
Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision. Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3: Organise paper definitions*). This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4: Organise uncertainty quantification*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 4: Organise uncertainty quantification*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.

Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues.

## Taxonomy definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7: Uncertainty Taxonomies*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

# Future work
This paper has identified several issues in the uncertainty visualisation literature and with those issues, we have several suggestions for future work that would likely be fruitful for the field.

Our first suggestion for future work is a mathematical definition of uncertainty. There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately. Elements such as imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between. A conceptual framework that allows us to combine these results would be incredibly beneficial to the field.

The concept of uncertainty should also be formalised within the grammar of graphics. This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation. Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results. incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.

Additionally the experimental practices within uncertainty visualisation need to be less ad hoc and more standardised. If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design. As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer. There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.

If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too. It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot. Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider. It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments. @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point. Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.

>>>>

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---

@@

# What is Uncertainty?
<!--## Colloquial definitions - what's in this section is not colloquial definitions, it's more overview-->

Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used. <!--is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. 
-->
Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain. Some works (XXX eg) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others (XXX eg)  <!--consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and-->  include broader loosely related elements, such as missing values (*Cite Gap 3: Organise paper definitions*). <!--This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4: Organise uncertainty quantification*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 4: Organise uncertainty quantification*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs. 
-->
Several attempts to address this gap are made in the papers describing taxonomies. 
<!-- Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues. -->

## Definitions in taxonomies
<!--Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.--> Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7: Uncertainty Taxonomies*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

# Future work
This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field. <!--and with those issues, we have several suggestions for future work that would likely be fruitful for the field.-->

<!--Our first suggestion for future work is--> *Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions. <!--There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately. Elements such as  imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between.--> A conceptual framework that allows us to combine these components is needed.

*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation. Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results. incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.

XXX Is data pipeline connected with the grammar of graphics? Should this be a recommendation?

*Experimental practices on uncertainty visualisation need to be standardised.* If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design. As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer. There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.

If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too. It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot. Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider. It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments. @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point. Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community. XXX Not sure what this paragraph is recommending?
